From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-05 Pi+/paper/reversible1.tex
intro

The circuit model of reversible computation discussed in the introduction is a useful abstraction close to the hardware
platform. However, since its main data abstraction is a \emph{sequence of wires}, it only provides an ``assembly''
programming abstraction (e.g., \verb|qasm|). A mathematical model based on permutations of finite sets provides a richer
algebraic model which we review in this section.

@@

A specialized language for programming with permutations on finite sets can be built using ideas going back to Mac
Lane~\cite{XXX} that the commutative semiring (also known as the commutative rig) of natural numbers exactly
characterizes permutations on finite sets. In programming parlance, as shown in Fig.~\ref{fig:pi-terms}, each primitive
algebraic identity of commutative rigs becomes a constant in the programming language and the algebraic closure
operators become composition operators in the programming language. Once equipped with a notion of values and types, we
get the following syntactic presentation of a programming language for
permutations~\cite{James:2012:IE:2103656.2103667,Carette2016}:

>>>>

The circuit model of reversible computation discussed in the introduction is a useful abstraction close to the hardware
platform. However, since its main data abstraction is a \emph{sequence of wires}, it only provides an ``assembly''
programming abstraction (e.g., \verb|qasm|). As motivated by \citet{LAFONT2003257}, a mathematical model based on
permutations of finite sets provides a richer algebraic model which we review in this section.

@@

A specialized language for programming with permutations on finite sets can be built using ideas going back to Kelly,
Laplaza, and Mac Lane~\cite{laplaza72,kelly74,KELLY197197} that the commutative semiring (also known as the commutative
rig) of natural numbers exactly characterizes permutations on finite sets. In programming parlance, as shown in
Fig.~\ref{fig:pi-terms}, each primitive algebraic identity of commutative rigs becomes a constant in the programming
language and the algebraic closure operators become composition operators in the programming language. Once equipped
with a notion of values and types, we get the following syntactic presentation of a programming language for
permutations~\cite{James:2012:IE:2103656.2103667,Carette2016}:
